{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-03-28 paper.tex
Case study part 1

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1em}
\caption{Correcting an incorrect formula. The user uses Copy to create the second
formula (1). They notice an error (2), go back in history to switch left and right argument (3),
merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

% TODO: This is potentially interesting too, but it is maybe way too complex and weird
%
% \subsubsection*{Dependency Tracking}
% Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
% conflict detection mechanism of Denicek (\S\ref{sec:system-ops}). When pushing edit $e_2$ through
% $e_1$, a conflict is detected if they affect overlapping targets or if $e_1$ modifies a
% document node that is a source of $e_2$, typically, the source of a \ident{Copy} edit.
%
% For edits generated during formula evaluation (\S\ref{sec:impl-eval}), we need to track an
% additional kind of dependency. Evaluating a formula that adds two numbers from two document
% location results in an edit that sets the result of the formula to the sum of the two numbers.
% The edit contains the resulting number, but we also need to record that the edit has the two
% source locations as dependencies. For this reason, Denicek makes it possible to annotate edits
% with additional dependencies. One alternative would be to introduce a special \ident{Evaluate}
% edit that would identify an operation to compute and a list of arguments, specified as a list
% of references. We choose the former to keep the set of edits smaller.

\subsubsection*{Merging Structure and Capabilities}
Denicek keeps a linear history and merging appends edits to the top of the history.
This model is akin to git rebase. One alternative is to maintain a graph of edits akin
to git merge. This would simplify recording of edits in programming by demonstration
(\S\ref{sec:impl-pbd}) as we would not need to update saved edits transformed
during merging. (Their hashes would remain the same.) However, supporting special
merge edits and non-linear history would make the basic substrate more complex.

Denicek also supports only the simpler \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
where users have to merge all changes in order and cannot merge only selected edits (cherry
picking in git). The \emph{divergence} model lets users, for example, keep their own schema but
import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
that is dual to our edit reconcilliation (given two subsequent edits, $e_1, e_2$, we want
to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).

\section{Case study}
\label{sec:case}
(data science environment)

>>>>

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1.5em}
\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

\subsubsection*{Merging Structure and Capabilities}
Denicek keeps a linear history and merging appends edits to the top of the history.
This model is akin to git rebase. An alternative is to maintain a graph of edits akin
to git merge. This would simplify recording of edits in programming by demonstration
(\S\ref{sec:impl-pbd}) as we would not need to update saved edits transformed
during merging. (Their hashes would remain the same.) However, supporting special
merge edits and non-linear history would make the basic substrate more complex.

Denicek also only supports the \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
where users have to merge all changes in order and cannot adopt selected edits (cherry
picking in git). In the \emph{divergence} model, users could keep their own schema but
import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, we want
to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).

\subsubsection*{Dependency Tracking}
Conflict detection in Denicek is used when merging edits, but also to implement
incremental recomputation (\S\ref{sec:impl-incremental}). Edits generated during formula evaluation
(\S\ref{sec:impl-eval}) may be \ident{Copy} edits (when evaluating a reference), but also
\ident{Add} edits (when setting the evaluation result to a computed value). For those edits, we
record additional dependencies identifying the source of the operation arguments. Those additional
dependencies also need to be considered in conflict detection.

% Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
% conflict detection mechanism of Denicek (\S\ref{sec:system-ops}). When pushing edit $e_2$ through
% $e_1$, a conflict is detected if they affect overlapping targets or if $e_1$ modifies a
% document node that is a source of $e_2$, typically, the source of a \ident{Copy} edit.
%
% For edits generated during formula evaluation (\S\ref{sec:impl-eval}), we need to track an
% additional kind of dependency. Evaluating a formula that adds two numbers from two document
% location results in an edit that sets the result of the formula to the sum of the two numbers.
% The edit contains the resulting number, but we also need to record that the edit has the two
% source locations as dependencies. For this reason, Denicek makes it possible to annotate edits
% with additional dependencies. One alternative would be to introduce a special \ident{Evaluate}
% edit that would identify an operation to compute and a list of arguments, specified as a list
% of references. We choose the former to keep the set of edits smaller.

\begin{figure}[t]
\vspace{-0.25em}
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
\vspace{-0.5em}
\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
notebook system.}
\label{fig:source}
\vspace{-0.25em}
\end{figure}

\begin{figure*}[t]
\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
a code cell (1), edits it in a grid that infers edit operations via programming by demonstration
(2), turned to a code cell (3) and then visualized.}
\label{fig:datnicek}
\end{figure*}

\section{Case study: Datnicek Notebooks}
\label{sec:case}

Denicek is intended as a low-level computational substrate that can be used as the basis for
interactive programming systems that view programs as documents. The most prominent example
of such systems today are notebook environments for data science. We developed Datnicek, a
simple notebook system that shows the ability of Denicek to support rich interactive user
experiences. In this section, we present Datnicek and reflect on its development. A simple data
exploration conducted in Datnicek is shown in Fig.~\ref{fig:datnicek}.

\subsection{Requirements}
The design of Datnicek brings together a range of recent research ideas on interactive programming
environments for data science. Datnicek notebooks consist of code cells and markdown cells, but
they also support interactive grid cells where the user can use programming by demonstration to
construct data cleaning scripts. We aim to support the following features:

\begin{itemize}
\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
  We do not aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.

\item \emph{Collaborative Editing.} It should be possible to merge independently done code changes
  as in Grove~\cite{adams-2025-grove}, addressing the known pitfall of versioning Jupyter
  notebooks~\cite{singer-2020-jollity}.

\item \emph{Code Completion.} During editing, code completion should offer available data
  transformations and operations as when using type providers \cite{syme-2013-inforich} or
  iterative prompting in The Gamma \cite{petricek-2022-thegamma}.

\item \emph{Output Invalidation.} When code is edited, the previously evaluated results that
  depend on it, directly or transitively, should be invalidated as in Wrattler \cite{petricek-2018-wrattler,petricek-2020-live},
  addressing another well-known limitation of Jupyter notebooks \cite{koop-2017-dataflow}.

\item \emph{Interactive Data Cleaning.} It should be possible to edit data tables in an interactive
  grid and use programming by demonstration for common data cleaning tasks as in
  Wrangler or Vizier~\cite{kandel-2011-wrangler,kennedy-2022-vizier}.

\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
  turned into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
\end{itemize}

\noindent
The Datnicek notebook system implements the above requirements on top of Denicek.
Although Datnicek is a proof of concept, it shows that Denicek provides a suitable basis
for the implementation.

\subsection{Implementation}

data representation

(maybe show fig with source)

operations

\newpage

\subsection{Reflections}

foof

(data science environment)
